Assessing the Impact of Vision Zero on Traffic Collisions in Toronto Neighborhoods (2014–2021):

### 2024-12-15 paper/paper.qmd
Paper Done!

---
title: "Assessing the Impact of Vision Zero on Traffic Collisions in Toronto Neighborhoods (2014–2021):"
subtitle: "An Unexpected Increase Following Policy Implementation"
author: 
  - Prankit Bhardwaj
  - Another author
thanks: "Code and data are available at: [https://github.com/prankitbhardwaj/Toronto-Traffic-Collisions]"
date: 3/12/2024
date-format: long
abstract: "This study analyzes traffic collision data from Toronto neighborhoods between 2014 and 2021 to evaluate factors influencing collision frequencies, with a focus on the impact of the Vision Zero Road Safety Plan implemented in 2017. Using a Negative Binomial regression model to account for overdispersion in count data, we found that while collision counts decreased annually over the study period, there was a significant increase in collisions following the implementation of Vision Zero. This unexpected result suggests that the policy has not yet achieved its intended effect of reducing traffic collisions. These findings highlight the need for a critical reassessment of road safety strategies and underscore the complexity of factors affecting traffic collisions in urban environments."
format: 
  pdf:
    toc: true
    toc_depth: 2
    number-sections: true
    include_in_header: preamble.tex
    latex-engine: xelatex
    include-in-header: header.tex
number-sections: true
bibliography: references.bib
---
\newpage

@@

library(tidyverse)
library(MASS)
library(ggplot2)
library(caret)
library(here)
library(broom) 
```

@@

Traffic collisions pose a significant public health and safety challenge worldwide, accounting for over 1.35 million deaths annually [@WHO2018]. In urban centers like Toronto, the complexities of traffic dynamics, urban planning, and population growth exacerbate the risks associated with road transportation. Understanding the patterns and determinants of traffic collisions is essential for developing effective interventions to enhance road safety.

Despite concerted efforts by city authorities, including the implementation of the Vision Zero Road Safety Plan in 2017, Toronto continues to grapple with high rates of traffic collisions, injuries, and fatalities [@CityOfToronto2017]. Previous studies have explored factors influencing collision rates, such as driver behavior, weather conditions, and infrastructure design [@Ma2019; @Wazana2020].. However, there remains a critical gap in comprehensively analyzing spatial and temporal trends at a granular neighborhood level, particularly in assessing the impact of policy interventions over time.

@@

1. **Traffic Collision Data**: Detailed records of reported traffic collisions in Toronto from January 2014 to December 2021, obtained from the **City of Toronto's Open Data Portal** (City of Toronto 2022a). The dataset includes information on collision dates, times, locations, severities, and parties involved.

@@

2. **Toronto Neighborhood Boundaries**: Geospatial data defining the boundaries of Toronto's 140 neighborhoods, sourced from the **City of Toronto's Open Data Portal** (City of Toronto 2022b).

@@

- **OCC_DATE**: Date and time of the collision occurrence (**POSIXct** format).
- **OCC_YEAR**: Year of occurrence (**integer**).
- **OCC_MONTH**: Month of occurrence (**factor** with levels "January" to "December").
- **OCC_DOW**: Day of the week (**factor** with levels "Monday" to "Sunday").
- **OCC_HOUR**: Hour of the day (**integer** from 0 to 23).
- **NEIGHBOURHOOD_NAME**: Name of the neighborhood where the collision occurred (**factor**).
- **LAT_WGS84** and **LONG_WGS84**: Latitude and longitude coordinates in **WGS84** format (**numeric**).
- **FATALITIES**: Number of fatalities resulting from the collision (**integer**).
- **INJURIES**: Number of injuries reported (**integer**).
- **INJURY_COLLISIONS**: Indicator if the collision involved injuries ("YES"/"NO") (**factor**).
- **AUTOMOBILE**, **MOTORCYCLE**, **CYCLIST**, **PEDESTRIAN**: Indicators for the types of road users involved ("YES"/"NO") (**factors**).

@@

- **NEIGHBOURHOOD_NAME**: Name of the neighborhood (**matches with collision data**).
- **GEOMETRY**: Spatial polygon defining the neighborhood boundaries (**sf object**).

@@

1. **Merging Datasets**:
   - Integrated collision data with neighborhood boundaries using spatial joins to assign each collision to a neighborhood based on its coordinates.

@@

```{r}
#| label: data-cleaning
#| fig-cap: Data Cleaning Process
#| echo: false
#| cache: TRUE
#| fig-pos: "H"

# Load necessary libraries
library(tidyverse)
library(lubridate)
library(sf)
library(here)

@@

# Convert OCC_DATE to POSIXct and adjust to EST
collisions <- collisions %>%
  mutate(
    OCC_DATE = as.POSIXct(OCC_DATE / 1000, origin = "1970-01-01", tz = "UTC"),
    OCC_DATE = with_tz(OCC_DATE, tzone = "America/Toronto"),
    OCC_YEAR = year(OCC_DATE),
    OCC_MONTH = month(OCC_DATE, label = TRUE, abbr = FALSE),
    OCC_DOW = wday(OCC_DATE, label = TRUE, abbr = FALSE),
    OCC_HOUR = hour(OCC_DATE)
  )

# Remove records with missing coordinates
collisions <- collisions %>%
  filter(!is.na(LAT_WGS84) & !is.na(LONG_WGS84))

@@

# Convert collisions to spatial data
collisions_sf <- st_as_sf(
  collisions,
  coords = c("LONG_WGS84", "LAT_WGS84"),
  crs = 4326,
  remove = FALSE
)

# Spatial join to assign neighborhoods
collisions_sf <- st_join(collisions_sf, neighbourhoods["AREA_NAME"])

# Clean up and rename columns
collisions_clean <- collisions_sf %>%
  rename(NEIGHBOURHOOD_NAME = AREA_NAME) %>%
  select(-geometry)

@@

@@

@@

We observed fluctuations in the total number of collisions over the years. Notably, there was a significant decrease in 2020, likely due to reduced traffic volumes during the COVID-19 pandemic lockdowns.

```{r}
#| label: fig-total-collisions
#| fig-cap: This line chart illustrates the annual number of traffic collisions reported in Toronto from 2014 to 2021
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 4
#| out.width: "60%"
#| out.height: "45%"

@@

Mapping collision frequencies reveals clusters of high collision densities in downtown and densely populated areas. High-density clusters are primarily located in downtown and other high-traffic areas, highlighting regions requiring targeted safety interventions.

```{r}
#| label: fig-collision-density
#| fig-cap: This map displays the density of traffic collisions across different neighborhoods in Toronto
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 4
#| out.width: "60%"
#| out.height: "45%"

@@

Analyzing the severity of collisions indicates that the majority result in property damage only, but a significant proportion involve injuries or fatalities.

```{r}
#| label: fig-collision-severity
#| fig-cap: This bar chart depicts the distribution of collision severities in Toronto
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 5.6
#| fig.height: 2.5
#| out.width: "60%"
#| out.height: "45%"

@@

Investigating how collision severity varies over time by illustrating fluctuations in fatal and injury-related collisions, providing insights into the effectiveness of road safety interventions over time.

```{r}
#| label: fig-severity-trends
#| fig-cap: This line graph shows the yearly trends in collision severity from 2014 to 2021
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 4
#| out.width: "60%"
#| out.height: "45%"

@@

We examined the involvement of pedestrians and cyclists in collisions.This line chart presents the annual number of collisions involving pedestrians and cyclists, as they are considered vulnerable, in Toronto from 2014 to 2021. The data highlights trends that can inform targeted safety measures for these groups.

```{r}
#| label: fig-vulnerable-users
#| fig-cap: Collisions Involving Vulnerable Road Users Over Time
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 2.8
#| out.width: "60%"
#| out.height: "45%"

@@

```{r model-results, echo=FALSE, message=FALSE}
#| label: tab:model-estimates
#| echo: false
#| message: false

@@

```{r residual-plot, fig.cap="Residuals vs. Fitted Values"}
#| echo: false
# Plot residuals vs. fitted values
residuals <- residuals(nb_model, type = "pearson")
fitted_values <- fitted(nb_model)

@@

```{r cross-validation}
# Set up cross-validation
set.seed(42)

@@

train_control <- trainControl(method = "cv", number = 5)

# Define the model formula
model_formula <- total_collisions ~ OCC_YEAR + PostVisionZero + NEIGHBOURHOOD_NAME + InjuryRate + FatalityRate

@@

The cross-validation results are summarized in **Table 2**.

@@

The estimated coefficients from the Negative Binomial regression model are summarized in **Table NB**.

@@

@@

To assess the model fit, we conducted a residual analysis by plotting the Pearson residuals against the fitted values  @ref(tab:residual-plot-1).

```{r residual-plot-2, fig.cap="Residuals vs. Fitted Values", echo=FALSE}
#| label: tab:residual-plot-1
# Plot residuals vs. fitted values
residuals <- residuals(nb_model, type = "pearson")
fitted_values <- fitted(nb_model)

@@

We performed 5-fold cross-validation to evaluate the predictive performance of the model. The results are summarized in **Table 2**.

@@

knitr::kable(cv_results, caption = "Table 2: Cross-Validation Performance Metrics")
```

**Table 2: Cross-Validation Performance Metrics**

@@

**Table 2: Cross-Validation Performance Metrics**

>>>>

---
title: "Assessing the Impact of Vision Zero on Traffic Collisions in Toronto Neighborhoods (2014–2021):"
subtitle: "An Unexpected Increase Following Policy Implementation"
author: 
  - Prankit Bhardwaj
thanks: "Code and data are available at: [https://github.com/prankitbhardwaj/Toronto-Traffic-Collisions]"
date: 12/3/2024
date-format: long
abstract: "This study analyzes traffic collision data from Toronto neighborhoods between 2014 and 2021 to evaluate factors influencing collision frequencies, with a focus on the impact of the Vision Zero Road Safety Plan implemented in 2017. Using a Negative Binomial regression model to account for overdispersion in count data, we found that while collision counts decreased annually over the study period, there was a significant increase in collisions following the implementation of Vision Zero. This unexpected result suggests that the policy has not yet achieved its intended effect of reducing traffic collisions. These findings highlight the need for a critical reassessment of road safety strategies and underscore the complexity of factors affecting traffic collisions in urban environments."
format: 
  pdf:
    toc: true
    toc_depth: 2
    number-sections: true
    include_in_header: preamble.tex
    latex-engine: xelatex
    include-in-header: header.tex
number-sections: true
bibliography: references.bib
---
\newpage

@@

library(tidyverse)
library(MASS)
library(ggplot2)
library(caret)
library(here)
library(broom) 
library(palmerpenguins)
```

@@

Traffic collisions pose a significant public health and safety challenge worldwide, accounting for over 1.35 million deaths annually [@WHO2018]. In major urban centers like Toronto, the complexities of increasing population density, complex transportation networks, and rapid urban development exacerbate the risks associated with road transportation. Understanding the spatial distribution and temporal progression of these collisions is critical for identifying risk factors and implementing interventions that enhance road safety.

Toronto’s policy environment provides a unique case for examining changes in collision dynamics over time. In 2017, the city launched the Vision Zero Road Safety Plan, aiming to eliminate traffic fatalities and serious injuries. Although various measures—ranging from infrastructure upgrades to public awareness campaigns—have been introduced, Toronto still experiences high rates of collisions, injuries, and fatalities [@CityOfToronto2017]. Previous research has explored influences on collision rates, including driver behavior, weather conditions, and roadway design [@Ma2019; @Wazana2020]. However, these analyses often do not delve into the granular neighborhood level or systematically assess how policy interventions affect localized patterns over time, leaving a critical gap in our understanding of Vision Zero’s impact at a fine spatial scale.

@@

1. **Traffic Collision Data**: Detailed records of reported traffic collisions in Toronto from January 2014 to December 2021, obtained from the **City of Toronto’s Open Data Portal** (City of Toronto 2022a). This dataset is produced by the Toronto Police Service for annual statistical reporting and public transparency, making it a comprehensive and official source for understanding collision occurrences. While other datasets (e.g., national databases, insurance claim records, or community-reported incidents) could potentially offer additional insights, these alternatives often lack detailed spatial resolution, consistent reporting standards, or the necessary temporal coverage that the Toronto dataset provides.

@@

2. **Toronto Neighborhood Boundaries**: Geospatial data defining the boundaries of Toronto's 140 neighborhoods, sourced from the **City of Toronto's Open Data Portal** (City of Toronto 2022b). These boundaries are used administratively and are stable over time, allowing for longitudinal neighborhood-level analysis.

@@

- **OCC_DATE**: Date and time of the collision occurrence (**POSIXct** format).
- **OCC_YEAR**: Year of occurrence (**integer**).
- **OCC_MONTH**: Month of occurrence (**factor** with levels "January" to "December").
- **OCC_DOW**: Day of the week (**factor** with levels "Monday" to "Sunday").
- **OCC_HOUR**: Hour of the day (**integer** from 0 to 23).
- **NEIGHBOURHOOD_NAME**: Name of the neighborhood where the collision occurred (**factor**).
- **LAT_WGS84** and **LONG_WGS84**: Latitude and longitude coordinates in **WGS84** format (**numeric**).
- **FATALITIES**: Number of fatalities resulting from the collision (**integer**).
- **INJURIES**: Number of injuries reported (**integer**).
- **INJURY_COLLISIONS**: Indicator if the collision involved injuries, providing a quick severity check ("YES"/"NO") (**factor**).
- **AUTOMOBILE**, **MOTORCYCLE**, **CYCLIST**, **PEDESTRIAN**: Indicators for the types of road users involved ("YES"/"NO") (**factors**). “YES” indicates that at least one party of that type was involved.

@@

- **NEIGHBOURHOOD_NAME**: Name of the neighborhood (**matches with collision data**).
- **GEOMETRY**: Spatial polygon that defines the neighborhood’s boundaries. These boundaries facilitate stable geographic comparisons over time (**sf object**).

@@

1. **Merging Datasets**:
   - Used a spatial join to assign collisions to neighborhoods. This step leverages the GEOMETRY of neighborhoods and the LAT/LONG of collisions.

@@

```{r}
#| label: data-cleaning
#| fig-cap: Data Cleaning Process
#| echo: false
#| message: false
#| warning: false
#| cache: TRUE
#| fig-pos: "H"

library(tidyverse)
library(lubridate)
library(sf)
library(here)

@@

# Convert OCC_DATE and filter missing coords
collisions <- collisions %>%
  mutate(
    OCC_DATE = as.POSIXct(OCC_DATE / 1000, origin = "1970-01-01", tz = "UTC"),
    OCC_DATE = with_tz(OCC_DATE, tzone = "America/Toronto"),
    OCC_YEAR = year(OCC_DATE),
    OCC_MONTH = month(OCC_DATE, label = TRUE, abbr = FALSE),
    OCC_DOW = wday(OCC_DATE, label = TRUE, abbr = FALSE),
    OCC_HOUR = hour(OCC_DATE)
  ) %>%
  filter(!is.na(LAT_WGS84) & !is.na(LONG_WGS84))

@@

# Spatial join
collisions_sf <- st_as_sf(
  collisions,
  coords = c("LONG_WGS84", "LAT_WGS84"),
  crs = 4326,
  remove = FALSE
) %>%
  st_join(neighbourhoods["AREA_NAME"])

collisions_clean <- collisions_sf %>%
  rename(NEIGHBOURHOOD_NAME = AREA_NAME) %>%
  dplyr::select(-geometry)

@@

## **Summary Statistics**

The table below provides a high-level snapshot of the dataset’s scale, temporal coverage, and spatial coverage. It illustrates the number of records, average fatalities per collision, the earliest and latest years of data, and how many neighborhoods are represented.

```{r}
#| label: tab:summary-statistics
#| fig-cap: Summary Statistics
#| echo: false
#| message: false
#| warning: false
#| cache: TRUE
#| fig-pos: "H"
summary_stats <- collisions_clean %>%
  summarize(
    tot_recs = n(),
    avg_fat = mean(FATALITIES, na.rm = TRUE),
    min_y = min(OCC_YEAR),
    max_y = max(OCC_YEAR),
    unique_neigh = n_distinct(NEIGHBOURHOOD_NAME)
  )

knitr::kable(summary_stats, caption = "Summary Statistics of Key Variables", booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("hold_position"))
```

@@

Below, we present a selection of visualizations to better understand distributions and patterns in the data. We focus on temporal patterns, spatial distributions, severity breakdowns, and involvement of vulnerable road users. While we refrain from interpretive explanations here, these visuals lay the foundation for insights discussed later.

@@

```{r}
#| label: fig-total-collisions
#| fig-cap: Annual number of traffic collisions reported in Toronto from 2014 to 2021
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 4
#| out.width: "60%"
#| out.height: "45%"

@@

```{r}
#| label: fig-collision-density
#| fig-cap: Map displaying total collisions per neighborhoods in Toronto
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 4
#| out.width: "60%"
#| out.height: "45%"

@@

```{r}
#| label: fig-collision-severity
#| fig-cap: Bar chart of collision severities in Toronto
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 5.6
#| fig.height: 2.5
#| out.width: "60%"
#| out.height: "45%"

@@

```{r}
#| label: fig-severity-trends
#| fig-cap: Yearly trends in collision severity(2014-2021)
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 4
#| out.width: "60%"
#| out.height: "45%"

@@

```{r}
#| label: fig-vulnerable-users
#| fig-cap: Annual Collisions Involving Vulnerable Road Users Over Time
#| echo: false
#| cache: TRUE
#| fig-pos: "H"
#| fig.width: 6
#| fig.height: 2.8
#| out.width: "60%"
#| out.height: "45%"

@@

```{r model-results}
#| label: model-estimates
#| echo: false
#| message: false
#| warning: false

@@

```{r residual-plot, fig.cap="Residuals vs. Fitted Values"}
#| echo: false

# Plot residuals vs. fitted values
residuals <- residuals(nb_model, type = "pearson")
fitted_values <- fitted(nb_model)

@@

```{r cross-validation}
#| cache: true
#| echo: false
#| message: false
#| warning: false

# Set up cross-validation
set.seed(42)

@@

train_control <- trainControl(method = "cv", number = 5, savePredictions = "final")

# Define the model formula
model_formula <- total_collisions ~ OCC_YEAR + PostVisionZero + NEIGHBOURHOOD_NAME + InjuryRate + FatalityRate
neighbourhood_yearly_collisions$NEIGHBOURHOOD_NAME <- droplevels(neighbourhood_yearly_collisions$NEIGHBOURHOOD_NAME)

@@

The cross-validation results are summarized in **Table 3**.

@@

The Negative Binomial model estimates are summarized below. This model includes year effects, a post-policy indicator (PostVisionZero), and severity-related rates. Neighborhood-specific effects are also included but not displayed here for brevity.

@@

### Coefficient Plot

The coefficient plot below shows point estimates and confidence intervals for the main predictors of interest, providing a visual representation of their direction and uncertainty.

```{r}
#| fig.cap: "Negative Binomial Model"
#| echo: false
#| message: false
#| warning: false
#| fig.width: 4
#| fig.height: 3.5
#| out.width: "45%"
#| out.height: "45%"
#| fig-align: "center"

coef_data <- broom::tidy(nb_model) %>%
  filter(term %in% c("OCC_YEAR","PostVisionZero","InjuryRate","FatalityRate"))

ggplot(coef_data, aes(x = term, y = estimate, ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error)) +
  geom_pointrange() +
  coord_flip() +
  labs(x = "Predictor",
       y = "Estimate (with 95% CI)") +
  theme_minimal()
```

@@

To assess the model fit, we conducted a residual analysis by plotting the Pearson residuals against the fitted values, plot belo.

```{r}
#| label: fig:residual-fitted-plot
#| echo: false
#| fig-cap: "Residuals vs. Fitted Values"
#| fig.width: 6
#| fig.height: 4
#| out.width: "60%"
#| out.height: "60%"
#| fig-align: "center"

# Plot residuals vs. fitted values
residuals <- residuals(nb_model, type = "pearson")
fitted_values <- fitted(nb_model)

@@

We performed 5-fold cross-validation to evaluate the predictive performance of the model. The results are summarized in **Table 5**.

@@

knitr::kable(cv_results, caption = "Cross-Validation Performance Metrics")
```

**Table 6: Cross-Validation Performance Metrics**

@@

**Table 7: Cross-Validation Performance Metrics**
